---
title: "ESM 244 Lab 1"
author: "Casey O'Hara"
date: "1/4/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf) # install.packages('sf')
library(tmap)
### install.packages('tmap')
### update.packages(ask = FALSE)
```

cmd-shift-enter shortcut for running the current code chunk

### Read in the data

cmd-option-i shortcut for creating a code chunk

```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'),
                     show_col_types = FALSE)
```

### Part 1: wrangling and ggplot review

**Example 1:** Find counts of observation by `legal_status` & wrangle a bit.

```{r}
### method 1: group_by() %>% summarize()
sf_trees %>%
  group_by(legal_status) %>%
  summarize(tree_count = n())

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>%
  count(legal_status) %>%
  drop_na(legal_status) %>%
  rename(tree_count = n) %>%
  relocate(tree_count) %>%
  slice_max(tree_count, n = 5) %>%
  arrange(desc(tree_count))
```

Make a graph of the top 5 from above:

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = 'Legal status', y = 'Tree count') +
  coord_flip() +
  theme_minimal()
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

shift-cmd-c to comment/uncomment quickly

```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>%
  filter(legal_status %in% c('Permitted Site', 'Private') & caretaker %in% 'MTA')
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude` and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>%
  filter(str_detect(species, 'Blackwood Acacia')) %>%
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = 'darkgreen')
```

**Example 4:** use `tidyr::separate()`

```{r}
sf_trees_sep <- sf_trees %>%
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

**Example 5:** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>%
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```




